---
title: 'POLS/CS&SS 503: Assignment 3'
author: Jeffrey Arnold
date: "May 17, 2017"
bibliography: assignment3.bib
output:
  html_document:
    toc: false
    number_sections: true
  pdf_document:
    toc: false
    number_sections: true
---

```{r include=FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
```


# Setup {-}

```{r}
library("tidyverse")
library("haven")
library("AER")
library("Formula")
```

# Problem 1

@Bailey2016a Ex 9.1

```{r}
RainIV <- read_csv( "Ch09.Ex1.CivilWarRainInstrumentData/RainIV.csv")
```

```{r}
ggplot(RainIV, aes(y = InternalConflict, x = LaggedGDPGrowth)) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r}
mod1_f <- InternalConflict ~ LaggedGDPGrowth
mod1 <- lm(mod1_f, data = RainIV)
summary(mod1)
```

**b.** Add controls for initial GDP (`InitialGDP`), democracy (`) mountains, and ethnic and religious fractionalization to the model.

```{r}
mod2_f <- update(mod1_f, . ~ . + InitialGDP + Democracy +
                 Mountains + EthnicFrac + ReligiousFrac)
mod2 <- lm(mod2_f, data = RainIV)
summary(mod2)
```

No. The coefficient on $\beta$ would only be a causal estimate under an assumption of no omitted confounders.
This seems highly unlikely.

**c.** The two conditions needed for a good instrument are 

- inclusion restriction: Rainfall must be correlated with economic growth
- exclusion restriction: Rainfall must not be correled with war excepct through its effect on economic growth.

The inclusion restriction can be tested (in the first stage):

```{r}
ggplot(RainIV, aes(x = LaggedRainfallGrowth, y = LaggedGDPGrowth)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The exclusion restriction cannot be tested statistically, and must be argued outside the model.

**d.** Instrumenting GDP growth with rainfall randomly assigns some part of GDP growth to countries.

**e.**
```{r}
mod3_f <- InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac | . - LaggedGDPGrowth + LaggedRainfallGrowth
mod3 <- ivreg(mod3_f, data = RainIV)
summary(mod3)
```
The coefficient on `LaggedGDPGrowth` is 20 times larger than 
in the OLS regression. It is still not stastistically significant.

f. Redo the 2SLS with country fixed effects.
```{r}
mod4_f <- InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_code | . - LaggedGDPGrowth + LaggedRainfallGrowth
mod4 <- ivreg(mod4_f, data = RainIV)
summary(mod4)
```

The coefficient on `LaggedGDPGrowth` has increased by 40% (-2 to -2.8) and now has a p-value of 0.06.
These regressions use only variation within each country and thus remove the country-level unobserved variables that may be correlated with economic growth. 


f. Regression `LaggedGDPGrowth` on rainfall and controls:
```{r}
mod_f1 <- lm(LaggedGDPGrowth ~ LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_code, data = RainIV)
```
Save the residuals. I use `broom::augment` for convenience, but still need to add `InternalConflict` since it wasn't in the original data.
```{r}
data_f <- broom::augment(mod_f1) %>% 
  left_join(select(RainIV, country_code, InternalConflict), by = "country_code")
```
```{r}
mod_f2 <- lm(InternalConflict ~ .resid + LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_code, data = data_f)
coef(mod_f2)["LaggedGDPGrowth"]
```

The coefficient is the same as that in the 2SLS.

It controls for endogeneity by controlling for the part of `LaggedGDPGrowth` not explained by rainfall growth (i.e. the non-random part of GDP growth).


# Problem 2

@Bailey2016a Ex 9.2

# Problem 3

@Bailey2016a Ex 9.4

# Problem 4

@Bailey2016a Ex 9.5

## References {-}
