---
title: 'POLS/CS&SS 503: Assignment 3'
author: Jeffrey Arnold
date: "May 17, 2017"
bibliography: assignment3.bib
output:
  html_document:
    toc: false
    number_sections: true
  pdf_document:
    toc: false
    number_sections: true
---

```{r include=FALSE}
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
```


# Setup {-}

```{r setup,message=FALSE}
library("tidyverse")
library("broom")
library("haven")
library("AER")
library("sandwich")
library("Formula")
```

# Problem 1

@Bailey2016a Ex 9.1

```{r}
RainIV <- read_csv( "Ch09.Ex1.CivilWarRainInstrumentData/RainIV.csv")
```

```{r}
ggplot(RainIV, aes(y = InternalConflict, x = LaggedGDPGrowth)) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r}
f_1a <- InternalConflict ~ LaggedGDPGrowth
mod_1a <- lm(f_1a, data = RainIV)
summary(mod_1a)
```

**b.** Add controls for initial GDP (`InitialGDP`), democracy (`) mountains, and ethnic and religious fractionalization to the model.

```{r}
f_1b <- update(f_1a, . ~ . + InitialGDP + Democracy + Mountains + EthnicFrac +
                 ReligiousFrac)
mod_1b <- lm(f_1b, data = RainIV)
summary(mod_1b)
```

No. The coefficient on $\beta$ would only be a causal estimate under an assumption of no omitted confounders.
This seems highly unlikely.

**c.** The two conditions needed for a good instrument are 

- inclusion restriction: Rainfall must be correlated with economic growth
- exclusion restriction: Rainfall must not be correled with war excepct through its effect on economic growth.

The inclusion restriction can be tested (in the first stage)
```{r}
ggplot(RainIV, aes(x = LaggedRainfallGrowth, y = LaggedGDPGrowth)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The exclusion restriction cannot be tested statistically, and must be argued outside the model.

**d.** Instrumenting GDP growth with rainfall randomly assigns some part of GDP growth to countries.

**e.**
```{r}
f_1e <- InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac | . - LaggedGDPGrowth + LaggedRainfallGrowth
mod_1e <- ivreg(f_1e, data = RainIV)
summary(mod_1e)
```
The coefficient on `LaggedGDPGrowth` is 20 times larger than 
in the OLS regression. It is still not stastistically significant.

f. Redo the 2SLS with country fixed effects.
```{r}
f_1f <- InternalConflict ~ LaggedGDPGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_code | . - LaggedGDPGrowth + LaggedRainfallGrowth
mod_1f <- ivreg(f_1f, data = RainIV)
summary(mod_1f)
```

The coefficient on `LaggedGDPGrowth` has increased by 40% (-2 to -2.8) and now has a p-value of 0.06.
These regressions use only variation within each country and thus remove the country-level unobserved variables that may be correlated with economic growth. 


g. Regression `LaggedGDPGrowth` on rainfall and controls:
```{r}
mod_1g <- lm(LaggedGDPGrowth ~ LaggedRainfallGrowth + InitialGDP + Democracy + Mountains + EthnicFrac + ReligiousFrac + country_code, data = RainIV)
```
Save the residuals. I use `broom::augment` for convenience, but still need to add `InternalConflict` since it wasn't in the original data.
```{r}
RainIV$resid <- residuals(mod_1g)
```
```{r}
mod_1g <- lm(InternalConflict ~ LaggedGDPGrowth + resid + InitialGDP +
               Democracy + Mountains + EthnicFrac + ReligiousFrac + country_code,
             data = RainIV)
coef(mod_1g)["LaggedGDPGrowth"]
```
The coefficient is the same as that in the 2SLS.
```{r}
coef(mod_1f)["LaggedGDPGrowth"]
```

It controls for endogeneity by controlling for the part of `LaggedGDPGrowth` not explained by rainfall growth (i.e. the non-random part of GDP growth).


# Problem 2

@Bailey2016a Ex 9.2

```{r}
NEWS_STUDY_FILE <- "Ch09.Ex2.TelevisionExperimentData/news_study_MAB.csv"
col_types <- cols(
  resid = col_character(),
  Female = col_character(),
  watchnat = col_character(),
  ReadNews = col_character(),
  pnintst = col_character(),
  Education = col_character(),
  income = col_character(),
  Voted = col_character(),
  prop = col_character(),
  infopro = col_character(),
  WatchProgram = col_character(),
  learnpro = col_character(),
  TreatmentGroup = col_character(),
  prop_dv = col_integer(),
  InformationLevel = col_integer(),
  prop_vote = col_integer(),
  white = col_integer(),
  partyid = col_integer(),
  PoliticalInterest = col_character()
)
news_study <- read_csv(NEWS_STUDY_FILE, col_types = col_types) %>%
  mutate(WatchProgram = if_else(WatchProgram == "yes", 1, 0),
         TreatmentGroup = if_else(WatchProgram == "0", 0, 1))
```

Estimate a regression with Proposition 209 as a dependent variable and whether the person watched the program as the independent variable,
```{r}
mod_news_study_a <- lm(InformationLevel ~ WatchProgram, data = news_study)
coeftest(mod_news_study_a, vcov. = vcovHC(mod_news_study_a))["WatchProgram", ]
```
Those who watched the television program on average report 0.3 higher points (p < .001) on the information about Proposition 209. 
This should not be interpreted causally, since it is almost certainly biased due to endogeneity.
One plausible example is that those more informed about politics are more likely to watch this TV program.

b. This regression controls for political interest, newspaper reading, and education.
```{r}
mod_news_study_b <-
  update(mod_news_study_a,
         . ~ . + PoliticalInterest + Education + ReadNews)
coeftest(mod_news_study_b, vcov. = vcovHC(mod_news_study_b))["WatchProgram", ]
```
The result is smaller 0.19 vs. 0.3 with a higher p-value (p < 0.05).
The identification strategy is selection on observables, which requires that all relevant variables are controlled for.
This is unlikely.

c. The assignment variable is a good instrument since it is an experiment, so by construction it is not correlated with the dependent variable and thus not associated with `PoliticalInterest` except watching the TV show.
```{r}
mod_news_study_c <- lm(WatchProgram ~ TreatmentGroup + PoliticalInterest +
                         Education + ReadNews, data = news_study)

coeftest(mod_news_study_c,
         vcov. = vcovHC(mod_news_study_c))["TreatmentGroup", ]
```


d. Estimate a 2SLS using `TreatmentGroup` as an instrument for `WatchProgram`:
```{r}
mod_news_study_d <-
  ivreg(InformationLevel ~ WatchProgram + PoliticalInterest +
          Education + ReadNews |
           . - WatchProgram + TreatmentGroup,
        data = news_study)

coeftest(mod_news_study_d)["WatchProgram", ]

coeftest(mod_news_study_d, vcov. = vcovHC(mod_news_study_d))["WatchProgram", ]
```


e. The 2SLS suggest that we can't reject that there is an effect of watching the program on information levels. Conditional on the IV assumptions (especially the inclusion restriction), this has removed endogeneity.


# Problem 3

@Bailey2016a Ex 9.4
```{r}
zipfile <- "Ch09.Ex2.TelevisionExperimentData.zip"
URL <- paste0("http://global.oup.com/",
              "us/companion.websites/fdscontent/uscompanion/us/",
            "static/companion.websites/9780199981946/data_sets/ch9/",
"Ch09.Ex4.EducationCrimeData.zip")
if (!file.exists(zipfile)) {
  download.file(URL, destfile = zipfile)
  unzip(zipfile) 
}
```


```{r}
EducationCrimeFile <- "Ch09.Ex4.EducationCrimeData/inmates.csv"
col_types <- cols(
  age = col_integer(),
  state = col_integer(),
  pob = col_integer(),
  gqtype = col_integer(),
  prison = col_integer(),
  educ = col_integer(),
  drop = col_integer(),
  AfAm = col_integer(),
  yearat14 = col_integer(),
  birthpl = col_integer(),
  req_sch = col_integer(),
  work_age = col_integer(),
  work_sch = col_integer(),
  ca = col_integer(),
  enrolage = col_integer(),
  drop_age = col_integer(),
  cl = col_integer(),
  ca8 = col_integer(),
  ca9 = col_integer(),
  ca10 = col_integer(),
  ca11 = col_integer(),
  cl6 = col_integer(),
  cl7 = col_integer(),
  cl8 = col_integer(),
  cl9 = col_integer(),
  year = col_integer(),
  hisp = col_character()
)
EducationCrime <- read_csv(EducationCrimeFile, col_types = col_types)
```


a. Run a LPM with prison as the dependent variable

```{r}
formula_ex3a <- prison ~ educ + age + AfAm + factor(year) + factor(state)
mod_ex3a <- lm(formula_ex3a, data = EducationCrime)
#coeftest(mod_ex3a, vcov. = vcovHC)["educ", ]
tidy(mod_ex3a) %>% filter(term == "educ")
```
On average those with one more year of schooling are 0.1% percent less likely to commit a crime, holding age and race constant (p < 0.001).

b. No. This depends on selection on observables. It is likely there is an unobservable that means that those who are more likely to complete schooling are less likely to commit crimes (SES, aggressiveness, conscientiousness).

c. Compulsary education laws should clearly be associated with more years of schooling, satisfying the inclusion restriction.
```{r}
mod_ex3c <- lm(educ ~ ca9 + ca10 + ca11 + age + AfAm + factor(state) +
                    factor(year), data = EducationCrime)
mod_ex3c_un <- lm(educ ~ age + AfAm + factor(state) + factor(year),
                     data = EducationCrime)
anova(mod_ex3c, mod_ex3c_un)
tidy(mod_ex3c) %>% filter(term %in% c("ca9", "ca10", "ca11"))

```
It's not as clear that it satisfies the exclusion restriction. 
It is plausible that states with more crime are more likely to pass compulsary schooling laws.

d. A 2SLS model using these instruments and robust se:
```{r}
mod_ex3d <- ivreg(prison ~ educ + age + AfAm + factor(state) + factor(year) |
                  . - educ + ca9 + ca10 + ca11, data = EducationCrime)
# coeftest(mod_ex3d, vcov. = vcovHC)
summary(mod_ex3d)
```



# Problem 4

@Bailey2016a Ex 9.5

```{r}
GrowthDemocracyFile <- "Ch09.Ex5.GrowthDemocracyData/democracy_income.csv"
col_types = cols(
  CountryCode = col_integer(),
  democracy_fh = col_double(),
  log_gdp = col_double(),
  year = col_integer(),
  worldincome = col_double(),
  YearOrder = col_integer()
)
GrowthDemocracy <- read_csv(GrowthDemocracyFile, col_types = col_types) %>%
  # add lag GDP
  group_by(CountryCode) %>%
  arrange(CountryCode, year) %>%
  mutate(lag_log_gdp = lag(log_gdp),
         lag_worldincome = lag(worldincome))
```

a. Run a model with Democracy as the dependent variable and logged GDP per capita as a 
```{r}
mod_ex5a <- lm(democracy_fh ~ lag_log_gdp, data = GrowthDemocracy)
tidy(mod_ex5a) %>% filter(term == "lag_log_gdp")
```

b. Include fixed effects of year and country in the previous model,
```{r}
mod_ex5b <- lm(democracy_fh ~ lag_log_gdp + factor(year) + factor(CountryCode),
   data = GrowthDemocracy)
tidy(mod_ex5b) %>% filter(term == "lag_log_gdp")
```
This uses only variation within country and within year. 
This explains much of the variation in democracy---e.g. time trends and variables
constant within country.

c. World income of trading partners should be associated with GDP. It is both
plausible, and has a t-stat of > 3 in the 1st stage regression.
```{r}
mod_ex5c <- lm(log_gdp ~ worldincome + factor(year) + factor(CountryCode),
               data = GrowthDemocracy)
tidy(mod_ex5c) %>% filter(term == "worldincome")
```
The exclusion restriction is not entirely plausible, but I can't think of a 
strong reason to object. Controlling for year is important since it will control
global trade shocks. What would be problematic is local changes which increase
both democracy and GDP per capita -- since most trading partners are close neighbors it
is plausible to think of spillovers. However, shocks to GDP in the short run seem
more plausible than shocks to democracy.

d. Run 2SLS with `worldincome` as an instrument for logged GDP,
```{r}
mod_ex5d <- ivreg(democracy_fh ~ lag_log_gdp +
                    factor(year) + factor(CountryCode) |
                  . - lag_log_gdp + lag_worldincome,
                  data = GrowthDemocracy)
summary(mod_ex5d)
```
The coefficient switches signs from 0.38 in the panel data to -0.21 in `ivreg` 
and the p-value decreases from 0.18 to 0.06.

It would probably be better to run this with cluster robust standard errors.

## References {-}
